Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-09-03 Manuscript/manuscriptPS.tex
Add author and update information about CHARMM36 simulations with different NBfix terms
Add new figures and text by Tiago Ferreira

\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
\author{Markus Miettinen}
\affiliation{Potsdam, Germany}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
\affiliation{Southampton, United Kingdom}
%\author{NMRlipids collaboration}
%\affiliation{nmrlipids.blogspot.fi}

@@

\subsection{Solid state NMR experiments}
The magnitude and signs of the C-H bond order parameters in
headgroup and glycerol backbone were measured
using natural abundance $^{13}$C solid state NMR spectroscopy as
described previously \cite{ferreira13,ferreira16}.
Shortly, the absolute values of the order parameters were determined from the dipolar splittings
given by the indirect dimension of 2D R-PDFL experiment \cite{dvinskikh04} and
the signs were measured using S-DROSS experiments \cite{gross97}.

\todo{Details of the used spectrometer and maybe some other details should be given.} \\
\todo{Sample preparation should be described.} \\
\todo{How is the peak assignment done?}

@@

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
carbons in headgroup and glycerol backbone region, except for g$_3$ for which the resolution
was not sufficient to determine the numerical value of the order paramater (Fig. \ref{PShgSIGNSsimpson}).
Slices of the R-PDFL spectra (Fig. \ref{PShgSIGNSsimpson} C)) 
show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
and a superposition of a large and a very small splitting for the $\alpha$-carbon.
The larger splitting gives a order parameter value of 0.09, while the numerical value
from the small splitting cannot resolved with the available resolution.
Since the R-PDFL and previous $^2$H NMR experiments \cite{browning80,roux91} give 
only the absolute values of order parameters, we determined the signs of PS headgroup
and glycerol backbone order parameters using the S-DROSS experiment \cite{gross97}.
The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
order parameter is negative, which is confirmed by SIMPSON simulations.
The beginning of the S-DROSS slice suggests that the higher order parameter of
the $\alpha$-carbon is positive and the deviation towards negative values with the longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
where the value of -0.02 was taken from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter. The S-DROSS curve from
SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} D)) did not agree with the experiment, 
confirming the interpretation that the smaller order parameter is negative.
\begin{figure}[!htb]
  \centering
  \includegraphics[width=8.5cm]{../scratch/figIDEA.eps}
  \caption{\label{PShgSIGNSsimpson}
    (a) The headgroup region of the INEPT spectrum with headgroup and glycerol backbone carbons assigned.
    (b) 2D R-PDLF spectra for headgroup and glycerol backbone regions.
    (c) Slices for $\alpha$ and $\beta$ barbons.
    (d) Experimental SDROSS data (points) and SIMPSON simulations (lines).
    Order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the larger and smaller $\alpha$-carbon slittings were used in the
    SIMPSON calculations. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller order parameter (dashed grey).
  }
  \todo{This is preliminary figure, should be polished.} \\
  \todo{Should we show slices for all the analyzed carbons in (c)?}
\end{figure}

>>>>

\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
\author{Markus Miettinen}
\affiliation{Potsdam, Germany}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
\affiliation{Southampton, United Kingdom}
%\author{NMRlipids collaboration}
%\affiliation{nmrlipids.blogspot.fi}

@@

\subsection{Solid state NMR experiments}

Headgroup and glycerol backbone C-H bond order parameters of POPS
were determined from chemical-shift resolved dipolar splittings
measured with a R-type Proton Detected Local Field (R-PDFL) experiment~\cite{dvinskikh04}
and corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
The following experimental setups were used.

{\emph{R-PDLF experiment}}. The parameters are described according to Figures 1c and 2c of the original reference
for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays $\tau_1$ and $\tau_2$ were 1.94 ms and 0.97 ms, respectively.
Radio frequency pulses with the nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
50 kHz (SPINAL64 $^1$H decoupling pulses). The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$ and 32 points in the indirect
dimension were recorded using 1024 scans for each, with recycle delay of 5 s and a spectral with of 149.5 ppm.

\emph{S-DROSS experiment}. The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
blocks $R$ were set as 39.4 $\mu$s and 89.4 $\mu$s, respectively. Radio frequency pulses with the nutation
frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H SPINAL64 decoupling).
The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s and a total of 8 points along $t_1$ were
measured using 1024 scans for each with a recycle delay of 5 s and a spectral with of 149.5 ppm.

\emph{NMR numerical simulations}
The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package using the $^{13}$C--$^1$H
dipolar couplings determined by the R-PDLF experiments or calculated from known $^2$H quadrupolar couplings as input.
Chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
distribution of bilayer orientations in the samples studied.

\emph{Sample preparation}
\todo{Sample preparation should be described.} \\

@@

\todo{How is the peak assignment done?}

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
carbons in headgroup and glycerol backbone region, except for g$_3$ for which the resolution
was not sufficient to determine the numerical value of the order paramater (Fig. \ref{PShgSIGNSsimpson}).
Slices of the R-PDFL spectra (Fig. \ref{PShgSIGNSsimpson} C)) 
show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
and a superposition of a large and a very small splitting for the $\alpha$-carbon.
The larger splitting gives a order parameter value of 0.09, while the numerical value
from the small splitting cannot resolved with the available resolution.
Since the R-PDFL and previous $^2$H NMR experiments \cite{browning80,roux91} give 
only the absolute values of order parameters, we determined the signs of PS headgroup
and glycerol backbone order parameters using the S-DROSS experiment \cite{gross97}.
The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
order parameter is negative, which is confirmed by SIMPSON simulations.
The beginning of the S-DROSS slice suggests that the higher order parameter of
the $\alpha$-carbon is positive and the deviation towards negative values with the longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
where the value of -0.02 was taken from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter. The S-DROSS curve from
SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} D)) did not agree with the experiment, 
confirming the interpretation that the smaller order parameter is negative.
\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller $\alpha$-carbon order parameter (dashed grey).
  }
\end{figure}

### 2018-09-03 Manuscript/manuscriptPSsuppl.tex
Add new figures and text by Tiago Ferreira

>>>>

\pagebreak
\section{Dipolar slices of the R-PDFL experiment}
\begin{figure}[]
%  \centering
  \includegraphics[width=20.0cm]{../Figs/SI_slices.pdf}
  \caption{\label{DPslices}
    Dipolar coupling slices from the R-PDFL experiment.
  }
  \todo{Labeling of carbons and explanations of numbers (chemical shift and order parameter) would be good.
  Also, the size of the figure in the file may not be optimal.}
\end{figure}
